\documentclass[11 pt, letterpaper]{report}

\usepackage[margin=1.25in]{geometry}
\usepackage{setspace}
\usepackage{enumitem}
\usepackage{csquotes}
\usepackage{amsmath}
\usepackage{float}
\usepackage{colortbl}
\SetBlockEnvironment{quotation}
%\usepackage[hidelinks]{hyperref}
\usepackage{url}
\usepackage{authblk}
\usepackage{adjustbox}
\usepackage{array}
\usepackage{fancyhdr}
\usepackage{svg}
\usepackage{multirow}
\usepackage{chngpage}
\doublespacing
\usepackage{graphicx}
\usepackage[format=plain,
            labelfont=bf]{caption}

\usepackage[sc]{mathpazo}

\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	linkcolor=black,
	citecolor=blue,
	urlcolor=blue
}

\usepackage{apacite}





\begin{document}
	
{\begin{titlepage}
		\large
		\singlespacing
		\begin{center}
			\mbox{}
			\vfill
			\textsc{Transportation, Poverty, and Urban Dynamics}\\
			\vfill
			by \\
			\vfill
			{Jeff Allen}\\
			\vfill
			\vfill
			A thesis submitted in conformity with the requirements \\
			for the degree of Doctor of Philosophy  \\
			Department of Geography and Planning \\
			University of Toronto \\
			\vfill
			{\copyright} Copyright 2022 by Jeff Allen\\
			\vspace{.01\textheight}
			\mbox{}
		\end{center}
		\setcounter{page}{1}
	\end{titlepage}
	\setcounter{page}{2}}






\renewcommand{\thepage}{\roman{page}}%

\newpage



\begin{center}
\section*{Abstract}
	\addcontentsline{toc}{section}{Abstract}
	\singlespacing
	{Title}\\[2ex]
	{Jeff Allen}\\
	{Doctor of Philosophy}\\
	Department of Geography and Planning\\
	University of Toronto\\
	{2022}\\
\end{center}

This dissertation examines suburbanization of poverty in Canadian cities during the late 20th and early 21st centuries, with particular focus on it's relation to the distribution of transportation infrastructure and travel behaviour outcomes. The body of this dissertation consists of a literature review followed by three quantitative research papers. The first examines suburbanization of poverty in Toronto over a 25 year period relative to changes in public transit accessibility and adverse travel behaviour outcomes (e.g. longer commute times). The second uses panel data to analyze and tabulate individual pathways to suburban poverty across Canada. And the third directly asks whether low-income residents are disproportionately moving away from public transit. Findings show that many suburban areas are not only declining in socio-economic status, but are also experiencing worsening travel restrictions, evidenced by longer commute times and lower activity participation rates. Importantly, the primary pathway to suburban poverty is sourced from residents dropping into poverty in the suburbs, rather than from moving away from central areas or due to immigration. Moreover, while low-income residents reduce their level of transit accessibility when they move, they are not doing so at a greater rate than higher-income movers. Overall, this research generates important knowledge about the changing structure of urban neighbourhoods while also providing pertinent information to aid preventative policy aimed at reducing suburban poverty in Canada.





\vspace{12mm}



\newpage


\chapter*{\Large{Acknowledgements}}

\addcontentsline{toc}{section}{Acknowledgements}

A PhD dissertation is a bit like a QQQQ. While there's only one name on the QQQQQ, it couldn't be accomplished without the help of many others.

The 21 pages of citations at the end of this document are clear evidence that uncountable words academics, journalists, bloggers, policy officials, and urban planners, among others who's words the ether of time and space, have influenced my thoughts and ideas.

Much of the empirical work in this dissertation was based on data collected and administrated by others. The Data Management Group and the Travel Modelling Group at the University of Toronto graciously provided the travel survey data and travel time matrices, respectively. The tax-record data was accessed at the Toronto Research Data Centre, and as such I would like to thank the staff at the RDC for their help in accessing and releasing the data. The data analysis throughout this dissertation was conducted via open source software (QGIS, Inksacpe, various R and Python libraries, etc.), and I am thankful to the open source community 

There are, importantly real people, too that . During this time, I am . . In particular, I would like to thank Nate Wessel, Mischa Young, Amber De John, and Matthew Palm for their (extra shout out to Matt who also provided excellent feedback on an earlier draft of Chapter 2).

My committee members, Chris Higgins and Dan Silver, 




Lastly, I'd like to thank my partner, Sarah Kaissi, and my mum, Carol Allen, for their unwavering love and support.


%
% R and Python programmers who create open source tools. QGIS communitiy.
%
% %RDC staff and data
%

% 
%
%Committee members
%
%Steve Farber. Money
%
%
%OGS scholarship, SSHRC scholarship. U of T moneys,
%
%Sarah, mum


\newpage





\tableofcontents

\addcontentsline{toc}{section}{Table of Contents}

\newpage






\addcontentsline{toc}{section}{List of Figures}

\listoffigures

\newpage


\addcontentsline{toc}{section}{List of Tables}

\listoftables



\hypersetup{
	colorlinks=true,
	linkcolor=red,
	citecolor=blue,
	urlcolor=blue
}



\newpage


\chapter*{Preface}

\addcontentsline{toc}{section}{Preface}

This dissertation can be read as a complete and coherent document, however Chapters \ref{ch:subtrapov}, \ref{ch:pathsubpov}, and \ref{ch:lowinctra} were originally written as separate articles to be published in peer-reviewed academic journals. As such, there is some redundancy in the introductory material, literature reviews, and methods sections of each. 

At the time of writing, Chapter \ref{ch:subtrapov} has been published in the Annals of the AAG \cite{allen_suburbanization_2021}; Chapter \ref{ch:pathsubpov} is under review at Environment and Planning A; and Chapter \ref{ch:lowinctra} is under review at the Journal of Transport and Land Use. I am first author on each of these publications and take full responsibility for omissions or errors in the analysis. 

As each of the chapters provides some of its own context, the goal of Chapter \ref{ch:intro} of this dissertation is to frame the issues motivating this research as a coherent body of work. Similarly, Chapter \ref{ch:conc} serves to tie together the results of each article into broader implications and research contributions as well as to suggest future lines of research. 



\newpage


\renewcommand{\thepage}{\arabic{page}}% Arabic numerals for page counter
\setcounter{page}{1}



\chapter{{Introduction}}

\input{chapters/1_intro.tex}
\label{ch:intro}



\chapter{Background}
\label{ch:background}
\input{chapters/2_background.tex}



\chapter{Suburbanization of Transport Poverty}
\label{ch:subtrapov}
\input{chapters/3_subtrapov.tex}



\chapter{Pathways to Suburban Poverty}
\label{ch:pathsubpov}
\input{chapters/4_pathsubpov.tex}



\chapter{Are low-income residents disproportionately moving away from transit?}
\label{ch:lowinctra}
\input{chapters/5_lowinctra.tex}




\chapter{Conclusions}
\label{ch:conc}
\input{chapters/6_conclusion.tex}


\bibliographystyle{apacite}
\bibliography{ref.bib}



%\nocite{*}


\end{document}